material and methods

### 2018-01-17 man/ms_1_material_methods.Rmd
update methods

### Greenness data

@@

EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016. A data filtering was applied to select EVI valid values. The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data. We filter out those values affected by high content of aerosols, clouds, shadows, snow or water.

Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2). Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked. Besides, VI usefulness indices between 0 to 11 essentially include all EVI data. Thus, these two conditions serve as additional checks.

@@

`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.

@@

OJITO---------- reescribir esto de arriba

After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2015. (:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano. Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md). Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)

@@

To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016]. For each pixel we averaged all the EVI valid values within a year (:red_filter: see quality filter), and then the standardized anomaly was computed as:

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.

Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1)

## Field sampling and dendrochronological methods

### Tree sampling 
Sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..). In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. We assessed stand competition of the target tree by recording all neighboring living trees with dbh > 7.5 cm in a circle of 10 m radi. The species, dbh and total height were annotated. We also measured distance and azimuth to the target tree. (:red_circle: indices competencia??)

@@

@@

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and mean, minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018, @Frias2018].
We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014]. SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017]. We select a temporal scale of 6 months.

@@

### Disturbance analyses

Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997]. ... This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
By using a temporal window of reasonable length, this method filters out the response to short-term changes in temperature and precipitation [@Nowacki1997, @Fraver2005].

@@

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median. We used medians since they are mores robust estimator of central tendency than means [@Rubino2004; @Camarero2011]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. A

To separate growth periods produced by disturbance events from those by climate, we considered a threshold of 25 % of GC for consi

A minimum $GC$ of 25% was established to depict canopy disturbances. Furthermore, more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.

This approach filters out

We used the percentage Growth-change filter (GC).

, calculated annually using 10-year windows

Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997). This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.

@@

Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.

The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990).

@@

 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]

* GC Nowacki??

@@

To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), recovery (*Rc*) and relative resilience (*RRs*).

The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:

@@

The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:

Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought

We computed the values of these indices for tree growth and greenness during each drought event. We considered 2005 and 2012 as singles drought events. The predrought and postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (:red_circle: incluir tabla de coeficientes y/o gráfica?? como suplement, see Gazol 2017)

@@

* Explore anomalies EVI 
* Explore long and short term trends in RW :red_circle: ver correo Guillermo
* ANOVA analysis EVI events and populations

We tested for significant differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the resilience indices. Robust two-way ANOVAs were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012, @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the WRS2 [@Mair2017] and rcompanion [@Mangiafico2017] R packages. The level of significance was set at 0.05 and adjusted for multiple comparisons.

All analyses were carried out in R software (ver. 2.11.1), using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA. All tests were at α = 0.05, except when indicated.

>>>>

## Greenness data

@@

MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). A data filtering was applied to select EVI valid values. The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data. We filter out those values affected by high content of aerosols, clouds, snow and shadows, following the filtering recomendations of EVI data for mountain regions [@Reyes2015].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2). Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked. Besides, VI usefulness indices between 0 to 11 essentially include all EVI data. Thus, these two conditions serve as additional checks.

@@

@@

")
```

After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016. EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).

```{r echo=FALSE, results='asis', eval=show_text}
cat("
(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano. Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md). Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)

@@

")
```

To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016]. For each pixel we annually averaged all the EVI valid values, and then the standardized anomaly was computed as:

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) (incluir cita). 
## Field sampling and dendrochronological methods

### Tree sampling 
Sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. We assessed stand competition of the target tree by recording all neighboring living trees with dbh > 7.5 cm in a circle of 10 m radi. The species, dbh and total height were annotated. We also measured distance and azimuth to the target tree. (:red_circle: indices competencia??)

@@

Statistical descriptive parameters [@Fritts1976] including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology.

@@

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and mean, minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014]. SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017]. We select a temporal scale of 6 months.

@@

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997]. By using a temporal window of reasonable length, this approach filters out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

@@

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median. We used medians since they are mores robust estimator of central tendency than means [@Rubino2004; @Camarero2011]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. A minimum $GC$ of 25% was established to depict canopy disturbances, and we considered a stand-wise disturbance when more than 50% of the individual trees display growth changes above 25 %.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997). This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

@@

@@

```{r echo=FALSE, results='asis', eval=show_text}
cat("
 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
")
```

@@

To evaluate the effects of the drougth events on greeennes and tree growth (BAI) we used three resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:

@@

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:

Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
")
```

We computed the values of these indices for tree growth (BAI) and greenness during each drought event. We considered 2005 and 2012 as singles drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (DUDA: incluir tabla de coeficientes y/o gráfica?? como suplement, see Gazol 2017).

@@

We tested for significant differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the resilience indices. Robust two-way ANOVAs were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set at 0.05 and adjusted for multiple comparisons.

All analyses were carried out in R software [@R2017] and  using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA. All tests were at α = 0.05, except when indicated.
